---
title: "Word prediction using ngrams"
author: "Francisco Javier Raya"
date: "3/30/2021"
output: ioslides_presentation
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction

I have developed a shiny app that predicts next word using and ngram model.

I have used a SwiftKey corpus. In this corpus I can find blogs, news and twitter data.

There are a lot of text, so I have reduced corpus size and filtering some words.

* Url of the shiny app: [https://fjraya.shinyapps.io/word_pred/]
* Url of github: [https://github.com/fjraya/final_ds_project]

## How have I cleaned the data?


* I have used quanteda library
* I have split text into sentences to not mix words from differents sentences
* I have removed punctuation, separators, convert all the text to lower case, filtering non ascii words and, finally, remove profanity words
* I haven't remove stop words because there are very present in language and is useful predict them and use them to predict more accurate.
* I have built ngrams models with 1, 2, 3 and 4 words. (see [https://en.wikipedia.org/wiki/N-gram] for more information)


## Algorithm used

* Probabilites are calculated with [Markov property](https://www.ee.ryerson.ca/~courses/ee8103/chap5.pdf) and [MLE estimation](https://leimao.github.io/blog/Maximum-Likelihood-Estimation-Ngram/).
* I have used [Katzs Backoff](https://en.wikipedia.org/wiki/Katz%27s_back-off_model) with precalculated [Good Turing discount](https://en.wikipedia.org/wiki/Good%E2%80%93Turing_frequency_estimation) stored in every ngram column.
* Using this algorithm, I manage unseen ngram words.
* When no ngram matches with input, I return most frequent word.
* It's important to say that all the process starts when I put 3 or more words in input text. For two or one word, this app doesn't give any prediction.



## Size and time reduction and evaluation

* I have deleted ngrams that occurs less than 2 times.
* I have studied word coverage and I only consider words that covers 90% of all frequencies.
* I have splitted corpus in train and test set (3000 trigrams) to evaluate different models with accuracy.

```{r table, echo=FALSE, message=FALSE, warning=FALSE}
library(data.table)
results <- data.table(model=c("baseline", "95% cover", "90% cover", "70% cover"), size=c("43", "29", "27", "20"), time=c("70", "9", "4", "2"), accuracy=c("0.21", "0.2", "0.199", "0.192"))

names(results) <- c("model", "size(Mb)", "time(min)", "accuracy")

head(results)
                      
                      
```

I have chosen model with 70% of coverage. Because is x2 faster and his size is lower. I think it's a good decision if I want to put this model in mobile or other devices with low resources. 


